On the Gravitational stability of gravito-turbulent accretion disks

### 2016-03-26 paper/application.tex
revised version done

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 2cm 0cm
    0.0cm]{figures/ppd_2d_growth}\\
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
    0.cm]{figures/ppd_2d_maxk}
  \caption{Black lines show growth timescales (top) of the most
    unstable wavenumber (bottom) for viscous, 
    self-gravitational modes in the 2D PPD model shown in
    Fig. \ref{rafikov_model}. Blue curves are for the same disk
    model but with a lower accretion rate. Solid curves are obtained
    numerically from Eq. \ref{thindisk}, and dotted curves 
    are analytic results based on Eq. \ref{gammie_smallk}. 
    {\bf 
      For both accretion rates the cooling time drops below $3/\Omega$
      beyond the vertical dashed line, but for the  
      $\dot{M}=10^{-6}M_\sun\mathrm{yr}^{-1}$ disk $\alpha > 0.1$
      beyond this radius, while $\alpha\lesssim
      0.03$  for the
      $\dot{M}=10^{-7}M_\sun\mathrm{yr}^{-1}$. 
    }
    \label{rafikov_growth}}
\end{figure}

>>>>

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 2cm 0cm
    0.0cm]{figures/ppd_2d_growth}\\
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
    0.cm]{figures/ppd_2d_maxk}
  \caption{Black lines show growth timescales (top) of the most
    unstable wavenumber (bottom) for viscous, 
    self-gravitational modes in the 2D PPD model shown in
    Fig. \ref{rafikov_model}. Blue curves are for the same disk
    model but with a lower accretion rate. Solid curves are obtained
    numerically from Eq. \ref{thindisk}, and dotted curves 
    are analytic results based on Eq. \ref{gammie_smallk}. 
    {\bf 
      For both accretion rates the cooling time $\lesssim 3/\Omega$
      beyond the vertical dashed line, but for the 
      $\dot{M}=10^{-6}M_\sun\mathrm{yr}^{-1}$ disk $\alpha > 0.1$
      beyond this radius, while $\alpha\lesssim
      0.03$ throughout the
      $\dot{M}=10^{-7}M_\sun\mathrm{yr}^{-1}$ disk. 
    }
    \label{rafikov_growth}}
\end{figure}

### 2016-03-26 paper/model3d.tex
revised version done

{\bf
Beta cooling of the form Eq. \ref{beta_cool} is widely applied in 2D
and 3D numerical simulations of self-gravitating disks \citep{gammie01,
  rice05,rice11,paardekooper12}. In fact, for $T_\mathrm{irr}=0$
Eq. \ref{beta_cool} the 
cooling function $\Lambda = -E/t_\mathrm{c}$ is identical
to that originally employed by \cite{gammie01}. 
We will refer to Eq. \ref{beta_cool} as  
`standard' beta cooling. It permits numerical experiments to be
carried out in a controlled manner as a function of the cooling time
$\beta$. An adiabatic disk corresponds to $\beta\to \infty$. The
physical meaning of the limit $\beta\to0$ depends on $T_\mathrm{irr}$, as
discussed in \S\ref{theta0}---\ref{theta1}.  
}

{\bf For standard beta cooling} we assume an 
equilibrium polytropic relation 
\begin{align}\label{poly_vert} 
  P  =
%  \frac{c_{s0}^2\rho_0}{\Gamma}\left(\frac{\rho}{\rho_0}\right)^\Gamma,
  %  \equiv \frac{c_s^2(z)}{\Gamma}\rho,
c_{s0}^2\rho_0\left(\frac{\rho}{\rho_0}\right)^\Gamma,
\end{align}
{\bf where $\rho_0 = \rho(z=0)$ is the equilibrium mid-plane density, and    
$\Gamma$ is the constant polytropic index which determines the disk's
  vertical structure. Thus $\Gamma$ is only relevant to the 3D
  problem.}     
The vertical structure is first obtained from 
Eq. \ref{vert_eq1}---\ref{vert_eq2}, %given strength of self-grav, via Q3d
then inserted into 
Eq. \ref{thermal_eq} to infer the required viscosity profile for
thermal equilibrium. If  $\mathcal{H}_\mathrm{ext}=0$, 
\begin{align}\label{alpha_beta_relation}
\alpha(z) = \frac{1}{(\gamma-1)\beta
   q^2}\frac{c_s^2(z)}{c_{s0}^2}\left(1 - \theta\right),
%  \alpha(z) = \frac{1}{\Gamma(\gamma-1)\beta
%    q^2}\left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} \left(1 - \theta\right). 
\end{align} 
{\bf with 
  \begin{align}\label{tirr_def}
    \theta = \frac{\tirr}{T_\mathrm{eq}},
  \end{align}
   where $T_\mathrm{eq}(z)$ is the equilibrium temperature field.
   We shall consider vertically isothermal disks with $\Gamma=1$, as
  appropriate for the outer parts of irradiated protoplanetary disks
   \citep{chiang97}. In this case  
   $\alpha$, $T_\mathrm{eq}$ and $\theta$ are simply constants. 
    Note that $\theta$  
    should only be interpreted as an irradiation parameter when it is 
    defined through Eq. \ref{tirr_def}, in conjunction with adopting  
    Eq. \ref{beta_cool} as the cooling function. 
 }

>>>>

{\bf
Beta cooling of the form Eq. \ref{beta_cool} is widely applied in 2D
and 3D numerical simulations of self-gravitating disks \citep{gammie01,
  rice05,rice11,paardekooper12}. In fact, for $T_\mathrm{irr}=0$ the 
cooling function $\Lambda = -E/t_\mathrm{c}$ is identical 
to that originally employed by \cite{gammie01}. 
We will refer to Eq. \ref{beta_cool} as  
`standard' beta cooling. It permits numerical experiments to be
carried out in a controlled manner as a function of the cooling time
$\beta$. An adiabatic disk corresponds to $\beta\to \infty$. The
physical meaning of the limit $\beta\to0$ depends on $T_\mathrm{irr}$, as
discussed in \S\ref{theta0}---\ref{theta1}.  
}

{\bf For standard beta cooling} we assume an 
equilibrium polytropic relation 
\begin{align}\label{poly_vert} 
  P  =
%  \frac{c_{s0}^2\rho_0}{\Gamma}\left(\frac{\rho}{\rho_0}\right)^\Gamma,
  %  \equiv \frac{c_s^2(z)}{\Gamma}\rho,
c_{s0}^2\rho_0\left(\frac{\rho}{\rho_0}\right)^\Gamma,
\end{align}
{\bf where $\rho_0 = \rho(z=0)$ is the equilibrium mid-plane density, and    
$\Gamma$ is the constant polytropic index that determines the disk's
  vertical structure. Thus $\Gamma$ is only relevant to the 3D
  problem.}     
The vertical structure is first obtained from 
Eq. \ref{vert_eq1}---\ref{vert_eq2}, %given strength of self-grav, via Q3d
then inserted into 
Eq. \ref{thermal_eq} to infer the required viscosity profile for
thermal equilibrium. If  $\mathcal{H}_\mathrm{ext}=0$, 
\begin{align}\label{alpha_beta_relation}
\alpha(z) = \frac{1}{(\gamma-1)\beta
   q^2}\frac{c_s^2(z)}{c_{s0}^2}\left(1 - \theta\right),
%  \alpha(z) = \frac{1}{\Gamma(\gamma-1)\beta
%    q^2}\left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} \left(1 - \theta\right). 
\end{align} 
{\bf with 
  \begin{align}\label{tirr_def}
    \theta = \frac{\tirr}{T_\mathrm{eq}},
  \end{align}
   where $T_\mathrm{eq}(z)$ is the equilibrium temperature field.
   We shall consider vertically isothermal disks with $\Gamma=1$, as
  appropriate for the outer parts of irradiated protoplanetary disks
   \citep{chiang97}. In this case  
   $\alpha$, $T_\mathrm{eq}$ and $\theta$ are simply constants. 
    Note that $\theta$  
    should only be interpreted as an irradiation parameter when it is 
    defined through Eq. \ref{tirr_def}, in conjunction with adopting  
    Eq. \ref{beta_cool} as the cooling function. 
 }

### 2016-03-26 paper/paper3d.tex
revised version done

{\bf 
  Global disk simulations \citep[e.g.][]{lodato04} show
  that the transport and 
  heating associated with this gravito-turbulence may be described as a
  local viscous process provided that the disk-to-star mass ratio is
  small ($\lesssim0.25$) and the disk is thin (aspect-ratio $\lesssim
  0.1$).  
}
{\bf In this case} the $\alpha$ and $\beta$ parameters defined above
will be inversely related (see, 
e.g. Eq. \ref{alpha_beta_relation}). {\bf Recent 
  vertically-extended shearing box simulations 
  also confirm this relation in 3D disks \citep{shi14}.}

@@

{\bf 
\subsection{Instability of the gravito-turbulent state}\label{gi_of_gi}
  In this work, we formally regard the gravito-turbulent
  disk described above as an equilibrium state, to which we apply    
  standard linear stability analysis. Thus by `perturbations' we mean   
  deviations away from this gravito-turbulent 
  basic state\footnote{This should not be confused with 
    flucuations with respect to the 
    laminar disk that maintain the underlying
    gravito-turbulence.}. 
  We assume instabilities defined this way  
  signifies disk fragmentation, since numerical simulations  
  show that the non-existance of steady gravito-turbulence imply    
  fragmentation.   
}

@@

%\acknowlegement
%S. Stahler 
%A. Youdin 
%paardekooper
%cathie clarke 
%lodato
%gammie

>>>>

{\bf 
  Global disk simulations \citep[e.g.][]{lodato04} show
  that the transport and 
  heating associated with this gravito-turbulence may be described as a
  local viscous process provided that the disk-to-star mass ratio is
  small ($\lesssim0.25$) and the disk is thin (aspect-ratio $\lesssim
  0.1$).  
}
{\bf In this case} the $\alpha$ and $\beta$ parameters defined above
{\bf are} inversely related (see, 
e.g. Eq. \ref{alpha_beta_relation}). {\bf Recent 
  vertically-extended shearing box simulations 
  also confirm this relation in 3D disks \citep{shi14}.}

@@

{\bf 
\subsection{Instability of the gravito-turbulent state}\label{gi_of_gi}
  In this work, we formally treat the gravito-turbulent
  disk described above as an equilibrium state, to which we apply    
  standard linear stability analysis. Thus by `perturbations' we mean   
  deviations away from this gravito-turbulent 
  basic state\footnote{This should not be confused with 
    fluctuations with respect to the 
    laminar disk that maintain the underlying
    gravito-turbulence.}. 
  We assume instabilities defined this way  
  signifies disk fragmentation, since numerical simulations  
  show that the non-existence of steady gravito-turbulence implies 
  fragmentation.   
}

@@

%\acknowlegement
{\bf
We thank the anonymous referee's prompt report that helped to improve
the clarity of this work. We thank S. Stahler and A. Youdin for
comments during the  course of this project; C. Clarke,
C. Gammie, and S.-J. Paardekooper for feedback on the first version of
this article, and G. Lodato for useful discussions.  
}
%S. Stahler 
%A. Youdin 
%paardekooper
%cathie clarke 
%lodato
%gammie

### 2016-03-26 paper/results2d.tex
revising
revised version done

{\bf
For example, we could assume that the viscosity only provides a
background heating and does not play an active role in the perturbed
state. This is in fact done implicitly in the literature when
discussing fragmentation or clump formation in cooling, 
self-gravitating disk simulations \citep{gammie01}. 
There, the effect of the ambient 
gravito-turbulent viscosity on the forming-clump is neglected, as 
one only compares adiabatic heating and the imposed cooling.  
This comparison is encapsulated in the generalized dispersion relation
below.}

@@

%for definiteness 
Let us consider a disk with 
\begin{align}\label{qcrit_def}
  \qtwo = \frac{1}{\sqrt{2\gamma(2-q)}}\equiv Q_{\mathrm{crit}},
\end{align} 
%{\bf KMK $\gamma = \theta$ from eq. 47? I'm missing something}
which is the condition for marginal stability in an adiabatic disk. 
How does finite cooling destabilize the disk?  
Inserting Eq. \ref{qcrit_def} into Eq. \ref{inviscid_max} with 
$\theta=0$, we find 
\begin{align}\label{sstar}
  S_*^3 = \frac{1}{\gamma Q_\mathrm{crit}^2 \beta}. 
\end{align}
The maximum growth rate, $S_*\propto \beta^{-1/3}$, smoothly
increases with decreasing $\beta$. 
%Instability formally occurs for
%arbitrarily long but finite cooling times. 
%Instability does not require the
%cooling time to be below some critical value.

>>>>

{\bf
For example, we could assume that the viscosity only provides a
background heating and does not play an active role in the perturbed
state. This is in fact done implicitly in the literature when
discussing fragmentation of cooling, 
self-gravitating disk simulations \citep{gammie01}. 
There, the effect of the ambient 
gravito-turbulent viscosity on the forming-clump is neglected, as 
one only compares adiabatic heating and the imposed cooling.  
This comparison is encapsulated in the generalized dispersion relation
below.}

@@

%for definiteness 
Let us consider a disk with 
\begin{align}\label{qcrit_def}
  \qtwo = \frac{1}{\sqrt{2\gamma(2-q)}}\equiv Q_{\mathrm{crit}},
\end{align} 
%{\bf KMK $\gamma = \theta$ from eq. 47? I'm missing something}
which is the condition for marginal stability in an adiabatic disk. 
How does finite cooling destabilize the disk?  
Inserting Eq. \ref{qcrit_def} into Eq. \ref{inviscid_max} with 
$\theta=0$, we find 
\begin{align}\label{sstar}
  S_*^3 = \frac{1}{\gamma Q_\mathrm{crit}^2 \beta}. 
\end{align}
The maximum growth rate, $S_*\propto \beta^{-1/3}$, smoothly
increases with decreasing $\beta$. {\bf Notice for $\beta>1$ this growth
  rate is faster than the imposed cooling rate $\beta^{-1}\Omega$}.
%Instability formally occurs for
%arbitrarily long but finite cooling times. 
%Instability does not require the
%cooling time to be below some critical value.

### 2016-03-26 paper/summary.tex
revising
revised version done

More recent simulations have raised the issue of numerical
convergence. %KMK 
\cite{meru11} found that better resolved disks fragmented at longer cooling times.
Follow-up studies attributed at least some of this effect to decreasing numerical viscous
heating at higher resolution, which helps fragmentation
\citep{lodato11,meru12}. 
We can expect this if numerical viscosity
contributes to an effective irradiation, %then we do expect a disk can be
%destabilized at longer cooling times with increasing resolution, 
because then perturbations can cool to lower temperatures with
increasing resolution, see Fig. \ref{invisc_theta}.

 \cite{paardekooper12} also found that fragmentation can occur 
for  slowly-cooled disks with $\tcool\Omega\gg O(1)$, but that this requires   
simulations to run for significantly longer than dynamical 
timescales. This is consistent with our finding
that for either cooling-driven or viscous GI, there is no critical 
cooling rate/viscosity below  which the disk is  
formally stable. Instead, growth rates smoothly decrease with
increasing $\tcool$ (decreasing $\alpha$), implying that
instabilities, and hence fragmentation, simply take longer to develop
for slowly-cooled disks.         
%hopkins paper

@@

Here, we highlight that most numerical experiments, including those
above, employ the standard beta cooling function, Eq. \ref{beta_cool},
\emph{without} a physical floor temperature. We show in \S\ref{2d_inviscid}
(see also \S\ref{cool_gi}) that this implies cooling-driven GI can
occur at any sufficiently small scale. Therefore as the numerical
resolution increases, simulations can access a wider range of unstable
scales. Although small-scale modes have weaker growth rates, they can 
become important over long timescales.  
%can access smaller, unstable scales 
In this respect, it is perhaps not surprising to find non-convergence with 
increasing resolution and/or integration times. 
%simple beta cooling will not give a
%well-imposed problem

@@

%KMK
A temperature floor may be a necessary, but not sufficient condition
for numerical convergence. \cite{baehr15} included a floor temperature 
in their simulations but still find that at fixed cooling rates, disks
eventually fragment with sufficient spatial resolution.
In light of our results on viscous GI, 
we suggest another possible contribution to non-convergence: %in addition to num visc (meru lodato)
high resolution enables small-scale turbulent angular momentum transport 
% physical, cf numerical  
to aid clump formation via the removal of rotational support. (See also \S\ref{MHD} below.) 
If clumps are only marginally resolved, simulations may not have sufficient dynamic range for turbulent eddies
to cascade down to these scales.
%This mechanism is unavailable at low resolutions, for which one must
%rely on rapid cooling.  
%maybe there is an effective floor temp which depends on res

>>>>

More recent simulations have raised the issue of numerical
convergence. %KMK 
\cite{meru11} found that better resolved disks fragmented at longer cooling times.
Follow-up studies attributed at least some of this effect to decreasing numerical viscous
heating at higher resolution, which helps fragmentation
\citep{lodato11,meru12}. 
We can expect this if numerical viscosity
contributes to an effective irradiation, %then we do expect a disk can be
%destabilized at longer cooling times with increasing resolution, 
because then perturbations can cool to lower temperatures with
increasing resolution, see Fig. \ref{invisc_theta}.  {\bf
  In global simulations, non-convergence has also been attributed to initial 
  conditions that lead to internal edges \citep{paardekooper11b}, but
  this cannot be modeled in our local setup. 
  \cite{rice12} point out that the standard implementation of beta cooling in
  SPH simulations apply cooling on scales well-below the SPH smoothing
  lengths, and that this inconsistency may contribute to
  non-convergence. 
}

{\bf However,} \cite{paardekooper12} also found {\bf in local 2D
  grid-based simulations} that fragmentation can occur  
for  slowly-cooled disks with $\tcool\Omega\gg O(1)$, but that this requires   
simulations to run for significantly longer than dynamical 
timescales. This is consistent with our finding
that for either cooling-driven or viscous GI, there is no critical 
cooling rate/viscosity below  which the disk is  
formally stable. Instead, growth rates smoothly decrease with
increasing $\tcool$ (decreasing $\alpha$), implying that
instabilities, and hence fragmentation, simply take longer to develop
for slowly-cooled disks.         
%hopkins paper

@@

Here, we highlight that most numerical experiments, including those
above, employ the standard beta cooling function, Eq. \ref{beta_cool},
\emph{without} a physical floor temperature. We show in \S\ref{2d_inviscid}
(see also \S\ref{cool_gi}) that this implies cooling-driven GI can
occur at any sufficiently small scale. Therefore as the numerical
resolution increases, simulations can access a wider range of unstable
scales. Although small-scale modes have weaker growth rates, they can 
become important over long timescales.   
%can access smaller, unstable scales 
In this respect, it is perhaps not surprising to find non-convergence with 
increasing resolution and/or integration times. 
{\bf
  On the other hand, the convergence issue may be
  less serious in 3D since small-scale modes are more stable in 3D
  than in 2D (Appendix \ref{3dcorr}). 
}
%simple beta cooling will not give a
%well-imposed problem

@@

%KMK
A temperature floor may be a necessary, but not sufficient condition
for numerical convergence. \cite{baehr15} included a floor temperature 
in their {\bf local 2D} simulations but still find that at fixed cooling rates, disks
eventually fragment with sufficient spatial resolution.
In light of our results on viscous GI, 
we suggest another possible contribution to non-convergence: %in addition to num visc (meru lodato)
high resolution enables small-scale turbulent angular momentum transport 
% physical, cf numerical  
to aid clump formation via the removal of rotational support. (See also \S\ref{MHD} below.) 
If clumps are only marginally resolved, simulations may not have sufficient dynamic range for turbulent eddies
to cascade down to these scales.
%This mechanism is unavailable at low resolutions, for which one must
%rely on rapid cooling.  
%maybe there is an effective floor temp which depends on res
